% Legal page — Harmonic Information Theory
\clearpage
\thispagestyle{empty}
\vspace*{0.025\textheight}
\begingroup
\setlength{\parskip}{0.22em}
\setlength{\parindent}{0pt}
\footnotesize
\raggedright
\leftskip=0.065\textwidth
\rightskip=0.065\textwidth

{\Large\bfseries Legal notice and license\par}
\vspace{0.22em}
\frontdivider[0.24\textwidth]

\vspace{0.4em}

{\frontlabel{Book and edition}\par}
\textbf{Harmonic Information Theory}\\
First digital edition\\
Jos\'e de la Quintana, C\'ordoba, Argentina --- 2026

\vspace{0.22em}
{\frontlabel{Credits}\par}
\textbf{Authors:} Mariano Fern\'andez M\'endez and Nicol\'as Ech\'aniz\\
\textbf{Compilation and final writing:} Mariano Fern\'andez M\'endez\\
\textbf{Institutional framework:} This work was developed and published within the institutional framework of \textbf{Asociaci\'on Civil AlterMundi}.\\
\textbf{Research and development team:} Anabella Scigliano Mattiauda, Federico Bonino, Javier Jorge, Juli\'an De La Reta, Mart\'in Fern\'andez M\'endez, Pablo Bustos, Saira Asua, Samanta La\~nin, Santiago Rodr\'iguez Cetr\'an, Sof\'ia Campagnoli.

\vspace{0.22em}
{\frontlabel{Publisher and legal address}\par}
\textbf{Publisher:} Asociaci\'on Civil AlterMundi\\
\textbf{Legal address:} A. Las Acacias 0, S:U205 M:68, Jos\'e de la Quintana, 5189, C\'ordoba, Argentina

\vspace{0.22em}
{\frontlabel{Official channels}\par}
\textbf{Official source:} \url{https://hit.altermundi.net}\\
\textbf{Contact:} \href{mailto:editorial@altermundi.net}{editorial@altermundi.net}

\vspace{0.22em}
\copyright\ 2026 Mariano Fern\'andez M\'endez and Nicol\'as Ech\'aniz

\vspace{0.22em}
{\frontlabel{License}\par}
Unless expressly indicated otherwise, the original content of this work is distributed under the license \textbf{Creative Commons Attribution 4.0 International (CC BY 4.0)} --- \url{https://creativecommons.org/licenses/by/4.0/}. You are free to copy, redistribute, share, adapt, remix, transform, and build upon this work in any medium or format, including for commercial purposes, provided that you give appropriate credit to the authors, provide a link to the license, indicate if changes were made, and do not suggest that the authors or Asociaci\'on Civil AlterMundi endorse your use or derivative version unless expressly authorized.

\vspace{0.22em}
{\frontlabel{Suggested attribution}\par}
Fern\'andez M\'endez, Mariano, and Nicol\'as Ech\'aniz. \emph{Harmonic Information Theory: Foundations}. Jos\'e de la Quintana, C\'ordoba, Argentina: AlterMundi, 2026. ISBN 978-631-91761-0-0. \url{https://hit.altermundi.net}. Compilation and final writing: Mariano Fern\'andez M\'endez. Institutional framework: Asociaci\'on Civil AlterMundi. License: CC BY 4.0.

\vspace{0.22em}
{\frontlabel{Publishing status}\par}
\textbf{ISBN:} 978-631-91761-0-0. This PDF is the \textbf{official bibliographic edition} of the book.

\vspace{0.22em}
{\frontlabel{Formats and official edition}\par}
The official bibliographic edition of the book is this \textbf{PDF}. The Markdown version and the LaTeX repository are part of the open publication and working ecosystem of the book, but do not constitute by themselves the bibliographic edition with ISBN.

\clearpage
\thispagestyle{empty}
\vspace*{0.025\textheight}
{\frontlabel{Third-party materials}\par}
Images, graphics, photographs, tables, extended quotations, or any other material identified as belonging to third parties are excluded from this license unless expressly indicated otherwise, and retain their respective terms of use.

\vspace{0.22em}
{\frontlabel{Special permissions, press, and institutional inquiries}: \href{mailto:editorial@altermundi.net}{editorial@altermundi.net}\par}

\vspace{0.22em}
{\frontlabel{Catalogaci\'on en la publicaci\'on (fuente)}\par}
{\scriptsize
Fern\'andez M\'endez, Mariano\\
\emph{Harmonic information theory : foundations} / Mariano Fern\'andez M\'endez ; Nicol\'as Ech\'aniz ; Contribuciones de Anabella Scigliano Mattiauda ... [et al.]. -- 1a ed. -- Jos\'e de la Quintana : AlterMundi, 2026.\\
Libro digital, PDF\\
Archivo Digital: descarga\\
Creada PARCIALMENTE con IA / Claude (Anthropic), Codex/OpenAI\\
ISBN 978-631-91761-0-0\\
1. Armon\'ia. 2. Filosof\'ia de la Informaci\'on. 3. Ciencias de la Informaci\'on. I. Scigliano Mattiauda, Anabella, colab. II. T\'itulo.\\
CDD 600\par}

\endgroup
